# Youth in the Canadian Labour Market
```{r}
#| label: lfs-setup

library(tidyverse)
source(here::here("R/import-lfs.R"))

d <- import_lfs_data()
```

## Definitions

### Youth

In this report, we use the term "youth" to refer to individuals of working age (15 to 29). 
This is consistent with definitions used by Statistics Canada and the International Labour Organization.

### Newcomer

This report defines newcomers as immigrants to Canada who landed 10 or fewer
years prior to the reference year. 
Conventional definitions of newcomers use a 5-year window, but the LFS PUMF does not
distinguish between immigrants who landed 5 or fewer years ago and those who landed 6 to 10 years ago.

### School Months

Current enrollment in either high school or post-secondary education is one dimension
of the segmentation model developed below. 
While the LFS PUMF variable `SCHOOLN` indicates current _attendance_, it does indicate current 
_enrollment_. 
This means that, for example, a student who is on summer break would be classified as not currently
attending school, even though they are still enrolled.
To address this issue, we limit impacted components of the analysis to typically active university terms (September to May). 

## Research Questions

1. How have youth fared in the Canadian labour market over the past 10 years?
  - Participation, employment, enrollment, and NEET rates 2015-2024
  - Average hourly wages 2015-2024
  - Job precarity indicators 2015-2024
2. How has the youth population changed?
  - Post-secondary attainment among youth 2015-2024
  - Representation of newcomers among youth 2015-2024
  - Segmentation model
3. How has youth search behaviour changed?
  - Job search rates among unemployed youth 2015-2024
  - Job search methods among unemployed youth by segment 2015-2024

## How have youth fared in the Canadian labour market?

Unsurprisingly, youth in Canada have lower participation, employment, and earnings than non-youth. 
Of note is the disproportionate impact of the COVID-19 pandemic on youth employment and participation rates in 2020, and their slower recovery compared to non-youth.
Whether this portends a long-term shift in the labour market remains to be seen, 
but it is clear that youth entering or returning to the labour market in the near future
will encounter a different landscape than their pre-pandemic counterparts. 

### Labour market rate indicators

@tbl-youth-rates presents annualized average rates of participation, employment, unemployment, enrollment, and NEET status among youth in Canada from 2015 to 2024. 

It suggests the following: 

- Youth participation and employment rates are consistently lower than those of the non-youth population.
- NEET rates among youth are consistently lower due to much higher rates of enrollment in education. 
- Youth participation, employment, and NEET rates jumped in 2020, but have largely recovered to pre-pandemic levels. 

```{r}
#| label: tbl-youth-rates
#| tbl-cap: "Labour market indicators for youth"
#| width: 100%

gt_spanners <- function(g, ...) {
  g_ <- g

  spanners <- rlang::list2(...)

  purrr::iwalk(
    spanners, 
    function(pattern, label) {
      g_ <<- g_ |> 
        gt::tab_spanner(
          label = label, 
          columns = matches(pattern)
        )
    }
  )
  g_
}

lfs_source_note <- function(g, u = "Working-age population (15-64)") {
  g |> 
    gt::tab_source_note(
      source_note = "Source: Labour Force Survey Public Use Microdata File (LFS PUMF), Statistics Canada"
    ) |> 
    gt::tab_source_note(
      source_note = glue::glue("Universe: {u}")
    )
}


t01 <- d |> 
  filter(school_month) |> 
  group_by(SURVYEAR, youth_fct) |> 
  summarize(
    participation = weighted.mean(participating_lgl, FINALWT),
    employment = weighted.mean(employed_lgl, FINALWT),
    unemployment = weighted.mean(unemployed_lgl, FINALWT, na.rm = TRUE),
    enrollment = weighted.mean(enrolled_lgl, FINALWT),
    neet = weighted.mean(neet_lgl, FINALWT)
  ) |> 
  ungroup()

t01 |> 
  pivot_wider(
    names_from = youth_fct, 
    values_from = c(participation, employment, unemployment, enrollment, neet)
  ) |> 
  gt::gt() |> 
  gt::fmt_percent(
    columns = c(-SURVYEAR),
    decimals = 0
  ) |> 
  gt_spanners(
    "Participation" = "participation", 
    "Employment" = "^employment", 
    "Unemployment" = "unemployment",
    "Enrollment" = "enrollment",
    "NEET" = "neet"
  ) |> 
  gtExtras::gt_theme_538() |> 
  gt::cols_label_with(
    columns = c(-SURVYEAR), 
    fn = function(x) {
      x |> 
        stringr::str_replace("^.*(A|Y).*(\\d{2}-\\d{2})$", "\\1(\\2)")
    }
  ) |> 
  lfs_source_note()
```


@fig-youth-rate-gap isolates the gap between youth and non-youth in participation, employment, and NEET rates. It highlights the disproportionate impact of the COVID-19 pandemic on youth employment and participation rates in 2020, and the slower recovery of youth employment and participation rates compared to those of non-youth. 

```{r}
#| label: fig-youth-rate-gap
#| caption: "Gaps between youth and non-youth in labour market indicators"

m01 <- d |> 
  filter(school_month) |>
  pivot_longer(
    c(participating_lgl, employed_lgl, neet_lgl),
    names_to = "var", 
    values_to = "y"
  ) |> 
  group_by(SURVYEAR, var) |> 
  group_nest(.key = "dat") |> 
  mutate(
    mod = map(
      dat,
      ~ lm(
        y ~ youth_lgl + factor(SURVMNTH), 
        data = .x,
        weights = FINALWT
      )
    ), 
    tab = map(
      mod, 
      ~ broom::tidy(.x, conf.int = TRUE) |> 
        dplyr::bind_rows(
          multcomp::glht(.x, "(Intercept) + youth_lgl = 0") |>
            broom::tidy() |>
            dplyr::select(
                -null.value
            ) |>
            dplyr::rename(
                term = contrast,
                p.value = adj.p.value
            ) |>
            dplyr::mutate(
                conf.low = estimate - 1.96 * std.error,
                conf.high = estimate + 1.96 * std.error
            )
        )
    )
  )

m01 |> 
  select(SURVYEAR, var, tab) |> 
  unnest(tab) |> 
  filter(term == "youth_lgl") |>
  ggplot(aes(SURVYEAR)) + 
  facet_wrap(~var) + 
  ggdist::stat_halfeye(
    aes(ydist = distributional::dist_normal(estimate, std.error))
  ) + 
  scale_x_continuous(
    breaks = 2015:2024, 
    labels = function(x) glue::glue("'{x - 2000}")
  ) + 
  scale_y_continuous(
    breaks = c(-.05, -.1, -.15, -.2),
    limits = c(-.2, -.05),
    labels = scales::percent_format(suffix = " ppt", accuracy = 1)
  ) + 
  labs(
    title = "Rate gaps between youth and non-youth",
    y = "Gap"
  )

```


### Job quality indicators

@tbl-youth-job-quality shows a selection of job quality indicators among employed youth in Canada from 2015 to 2024.

In it, we observe the following:
- Hourly earnings among employed youth are consistently lower than those of their non-youth counterparts.
- Employed youth are much more likely to occupy part-time or temporary positions than non-youth employees.
- Involuntary part-time work is more common among employed youth than among non-youth employees. 

```{r}
#| label: tbl-youth-job-quality
#| tbl-cap: "Job quality indicators for employed youth"

t02 <- d |> 
  filter(
    employed_lgl %in% 1, 
    as.integer(COWMAIN) < 3
  ) |> 
  group_by(SURVYEAR, youth_fct) |> 
  summarize(
    hrlyearn = weighted.mean(HRLYEARN, FINALWT),
    part_time = weighted.mean(FTPTMAIN == "Part-time", FINALWT),
    involuntary_pt = weighted.mean(job_involuntary_pt_lgl, FINALWT),
    temporary = weighted.mean(job_temporary_lgl, FINALWT), 
    .groups = "drop"
  )


t02 |> 
  pivot_wider(
    names_from = youth_fct, 
    values_from = c(-SURVYEAR, -youth_fct)
  ) |> 
  gt::gt() |> 
  gt::fmt_percent(
    columns = c(-SURVYEAR, -matches("hrlyearn")),
    decimals = 0
  ) |> 
  gt::fmt_currency(
    columns = matches("hrlyearn"),
    currency = "CAD",
    decimals = 2
  ) |>
  gt_spanners(
    "Hourly earnings" = "hrlyearn", 
    "Part-time" = "part_time",
    "Involuntary part-time" = "involuntary_pt",
    "Temporary" = "temporary"
  ) |> 
  gtExtras::gt_theme_538() |> 
  gt::cols_label_with(
    columns = c(-SURVYEAR), 
    fn = function(x) {
      x |> 
        stringr::str_replace("^.*(A|Y).*(\\d{2}-\\d{2})$", "\\1(\\2)")
    }
  ) |> 
  lfs_source_note(u = "Employed, employees")

```

@fig-youth-earnings-gap shows the gap in hourly earnings between youth and non-youth employees in Canada from 2015 to 2024. It suggests that earnings among employed youth have followed a similar trajectory to employment rates, in that they dropped disproportionately in 2020 and have lagged behind the recovery of non-youth earnings.

```{r}
#| label: fig-youth-earnings-gap
#| caption: "Earnings gap among youth"

m02 <- d |> 
  group_by(SURVYEAR) |> 
  group_nest(.key = "dat") |> 
  mutate(
    mod = map(
      dat,
      ~ lm(
        z_hrlyearn ~ youth_lgl + factor(SURVMNTH), 
        data = .x,
        weights = FINALWT
      )
    ), 
    tab = map(
      mod, 
      ~broom::tidy(.x, conf.int = TRUE) |>
        bind_rows(
          multcomp::glht(.x, "(Intercept) + youth_lgl = 0") |>
            broom::tidy() |>
            select(
                -null.value
            ) |>
            rename(
                term = contrast,
                p.value = adj.p.value
            ) |>
            mutate(
                conf.low = estimate - 1.96 * std.error,
                conf.high = estimate + 1.96 * std.error
            )
      )
    )
  )


m02 |> 
  select(SURVYEAR, tab) |> 
  unnest(tab) |> 
  filter(term == "youth_lgl")|>
  mutate(
    estimate = estimate * sd(d$HRLYEARN, na.rm = TRUE), 
    std.error = std.error * sd(d$HRLYEARN, na.rm = TRUE)
  ) |> 
  ggplot(aes(SURVYEAR)) + 
  ggdist::stat_halfeye(
    aes(ydist = distributional::dist_normal(estimate, std.error))
  ) + 
  scale_y_continuous(
    labels = scales::dollar_format()
  ) + 
  scale_x_continuous(
    breaks = 2015:2024
  ) + 
  labs(
    y = "Wage gap\n(naive likelihood)",
    title = "Youth wage disadvantage"
  ) + 
  bptheme::theme_blueprint(plot_background_color = NA)
```

## How has the youth population changed?

The youth population in Canada has grown more educated and more diverse over the past decade. 
The rate of PSE attainment increased by 9.6%, while the rate of newcomer representation among youth increased by 35.2% over the same period.

### National representation of PSE and newcomer status

@tbl-youth-pop shows the proportion of youth in Canada who have attained post-secondary education and the proportion of youth who are newcomers from 2015 to 2024.
Both indicators have increased over the past decade. 
The rate of post-secondary attainment among youth increased by 3.8 ppt between 2015 and 2024,
while the rate of newcomer representation among youth has increased by 3.0 ppt over the same period.  

```{r}
#| label: tbl-youth-pop
#| tbl-cap: "Youth PSE attainment and newcomer representation"

t03 <- d |> 
  filter(youth_lgl == 1, school_month) |> 
  group_by(SURVYEAR) |> 
  summarize(
    pse = weighted.mean(has_pse_lgl, FINALWT),
    newcomer = weighted.mean(newcomer_lgl, FINALWT)
  )

t03 |> 
  gt::gt() |> 
  gt::fmt_percent(
    columns = c(-SURVYEAR), 
    decimals = 1 
  ) |> 
  gtExtras::gt_theme_538() |> 
  gt::cols_label_with( 
    fn = function(x) {
      case_when(
        x == "pse" ~ gt::md("PSE attainment"),
        x == "newcomer" ~ gt::md("Newcomer representation"),
        x == "SURVYEAR" ~ gt::md("Year")
      )
    }
  ) |> 
  lfs_source_note(u = "Youth 15-29")
```

### Key segments of the youth population

@fig-youth-segments-rep shows the distribution of youth in Canada across key segments from 2015 to 2024.
The segments are defined by the intersection of educational enrollment, educational attainment, and newcomer status. 

```{r}
#| label: fig-youth-segments-rep
#| caption: "Segment representation among youth"
#| width: 100%
#| height: 3

t04 <- d |> 
  filter(
    youth_lgl == 1,
    school_month
  ) |> 
  group_by(SURVYEAR, SURVMNTH) |> 
  count(segment, wt = FINALWT) |>
  mutate(p = n / sum(n)) |>
  group_by(segment) |>
  mutate(idx_15 = round(p / first(p) * 100) - 100) |>
  ungroup()

t04 |> 
  ggplot(aes(SURVMNTH, p)) +
  facet_grid(.~SURVYEAR) +  
  geom_path(aes(group = segment, color = segment)) + 
  scale_x_discrete(
    labels = function(x) stringr::str_sub(x, end = 1L), expand = expansion(c(0.01, 0.01))
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  bptheme::theme_blueprint(plot_background_color = NA)
```

@fig-youth-segments-idx shows the percentage change in the representation of each segment 
in the youth population from January 2015. 
It highlights the dramatic proportional increase in the representation of non-student newcomers,
especially those with PSE attainment (which has more than doubled) among Canadian youth. 

```{r}
#| label: fig-youth-segments-idx
#| caption: "Shifts in segment representation among youth"
#| width: 100%
#| height: 3
t04 |> 
  ggplot(aes(SURVMNTH, idx_15)) +
  facet_grid(.~SURVYEAR) +  
  geom_path(aes(group = segment, color = segment)) + 
  scale_x_discrete(
    labels = function(x) stringr::str_sub(x, end = 1L), expand = expansion(c(0.01, 0.01))
  ) +
  scale_y_continuous() +
  bptheme::theme_blueprint(plot_background_color = NA)

```

@fig-youth-segments-prov disaggregates the total shift in segment representation from January 2015 to December 2024 by province. 

In absolute terms, representation of newcomers with PSE among working-age youth
increased in all provinces.
PEI experienced the highest absolute increase (+3.4 ppt), 
while Saskatchewan experienced the lowest (+.5 ppt).

```{r}
#| label: fig-youth-segments-prov
#| caption: "Shifts in segment representation among youth by province"
#| width: 100%
#| height: 6

t05 <- d |> 
  filter(youth_lgl == 1, school_month) |> 
  group_by(SURVYEAR, PROV) |> 
  count(segment, wt = FINALWT) |> 
  mutate(p = n / sum(n)) |>
  group_by(PROV, segment) |> 
  summarize(
    delta_abs = last(p) - first(p),
    delta_idx = last(p) / first(p) - 1,
    .groups = "drop"
  )

t05 |> 
  ggplot(aes(delta_abs, PROV)) + 
  geom_col(aes(fill = segment), position = "dodge") +
  facet_wrap(~segment, nrow = 2) + 
  bptheme::theme_blueprint(plot_background_color = NA) + 
  scale_x_continuous(labels = scales::percent_format(accuracy = 1), breaks = seq(-.03, .03, .01))
```

Indexed to 2015 (@fig-youth-segments-prov-idx), the growth of both newcomer segments dwarfs 
changes in the others. 
For those segments, changes were most dramatic in the Atlantic provinces.
Newfoundland and Labrador experienced the greatest relative increase, 
with the proportion of newcomers with PSE among working-age youth increasing
by an order of magnitude (12.6 X)

```{r}
#| label: fig-youth-segments-prov-idx
#| caption: "Shifts in segment representation among youth by province (indexed to 2015)"
#| width: 100%
#| height: 6

t05 |> 
  ggplot(aes(delta_idx, PROV)) + 
  geom_col(aes(fill = segment), position = "dodge") +
  facet_wrap(~segment, nrow = 2) + 
  bptheme::theme_blueprint(plot_background_color = NA) +
  scale_x_continuous(
    labels = function(x) {
      scales::number(x + 1, suffix = "x")
    }
  )
```

## How has youth job-search behaviour changed?

- Unemployed youth and adults behave similarly with respect to job search activities
- Among unemployed youth, newcomers with PSE---the fastest growing segment of the youth population---appear to have become the most active job seekers. 
Of the segments described above, they are the most likely to be searching for work, and the most likely to use public agencies, their personal networks, and job ads in their job search. 

### Compared with non-youth

- Unemployed adults 30+ and youth under 30 share similar job search behaviours
- Over time, those behaviours have trended in similar directions
- The only notable difference is that youth have been more likely to reach out to employers directly

```{r}
#| label: tbl-youth-search
#| tbl-cap: "Job search behaviour among unemployed youth"
#| width: 100%

t06 <- d |> 
  filter(unemployed_lgl == 1) |> 
  group_by(SURVYEAR, youth_fct) |> 
  summarize(
    job_searching = weighted.mean(job_searching_lgl, FINALWT), 
    across(
      matches("LK(PUBAG|EMPLOY|RELS|ATADS)"), 
      ~ weighted.mean(!is.na(.x), FINALWT, na.rm = TRUE)
    ), 
    .groups = "drop"
  )

t06 |> 
  pivot_wider(
    names_from = youth_fct, 
    values_from = c(-SURVYEAR, -youth_fct)
  ) |> 
  gt::gt() |> 
  gt::fmt_percent(
    columns = c(-SURVYEAR),
    decimals = 0
  ) |>
  gt_spanners(
    "Searched" = "job_searching", 
    "Public agency" = "LKPUBAG", 
    "Employer outreach" = "LKEMPLOY", 
    "Friends or family" = "LKRELS", 
    "Looked at ads" = "LKATADS"
  ) |> 
  gtExtras::gt_theme_538() |> 
  gt::cols_label_with(
    columns = c(-SURVYEAR), 
    fn = function(x) {
      x |> 
        stringr::str_replace("^.*(A|Y).*(\\d{2}-\\d{2})$", "\\1(\\2)")
    }
  ) |> 
  lfs_source_note(u = "Unemployed youth 15-30")
```

### By segment

- Among unemployed youth, newcomers _and_ those who are enrolled in or have completed PSE are more likely to rely on friends and family, 
especially in 2024
- Newcomers with PSE were the most likely segment to use public agencies, followed by newcomers without PSE
- Non-student newcomers with PSE---the fastest growing segment of the youth population---were, in 2024, the most likely segment to use public agencies, plumb their personal networks, and look at job ads in their job search


```{r}
#| label: tbl-segment-search
#| tbl-cap: "Job search behaviour by youth segment"
#| width: 100%

t07 <- d |> 
  filter(unemployed_lgl == 1, school_month, youth_lgl == 1) |> 
  group_by(SURVYEAR, segment) |> 
  summarize(
    job_searching = weighted.mean(job_searching_lgl, FINALWT), 
    across(
      matches("LK(PUBAG|RELS|ATADS)"), 
      ~ weighted.mean(!is.na(.x), FINALWT, na.rm = TRUE)
    ), 
    .groups = "drop"
  )

t07 |>
  filter(SURVYEAR %in% c(2015, 2020, 2024)) |> 
  pivot_wider(names_from = SURVYEAR, values_from = c(-SURVYEAR, -segment)) |>
  gt::gt() |> 
  gt::fmt_percent(
    columns = c(-segment),
    decimals = 0
  ) |>
  gt_spanners(
    "Searched" = "job_searching", 
    "Public agency" = "LKPUBAG", 
    "Friends and Family" = "LKRELS",
    "Looked at ads" = "LKATADS"
  ) |> 
  gtExtras::gt_theme_538() |> 
  gt::cols_label_with(
    columns = c(-segment), 
    fn = function(x) {
      x |> 
        stringr::str_replace("^.*(\\d{2})(\\d{2})$", "'\\2")
    }
  ) |> 
  lfs_source_note(u = "Unemployed youth 15-30") |> 
  gt::data_color(
    columns = c(matches("job_searching")), 
    palette = "Reds", 
    domain = c(0, 1)
  ) |> 
  gt::data_color(
    columns = c(matches("LKPUBAG")), 
    palette = "Blues", 
    domain = c(0, 1)
  ) |> 
  gt::data_color(
    columns = c(matches("LKRELS")), 
    palette = "Greens", 
    domain = c(0, 1)
  ) |> 
  gt::data_color(
    columns = c(matches("LKATADS")), 
    palette = "Purples", 
    domain = c(0, 1)
  )


```